Reconstructing the Lensing Mass in the Universe \\ from Photometric Catalogue Data

### 2012-11-22 doc/paper-lcr.tex
third draft finished

High precision cosmological distance measurements towards individual objects
such as time delay gravitational lenses or type Ia supernovae are affected by
weak lensing perturbations by galaxies and groups along the line of sight.
Such massive structures contribute an ``external convergence'' $\kappax$ that
in time delay gravitational lenses can dominate the uncertainty in the inferred distances.  We
characterise this uncertainty by marginalising over a probability density
function for $\kappax$: here, we investigate the use of a simple halo model to
estimate $\pr(\kappax)$ given noisy estimates of the photometric redshift
and stellar mass of galaxies observed in optical imaging surveys of the field
in question. We use mock catalogues from the Millennium Simulation, and compare
our reconstructed $\pr(\kappax|\data)$ to ray-traced $\kappax$ truth values: 
taking into account realistic uncertainties on redshift and stellar masses, we
typically find that the reconstruction yields a factor of 2 improvement in precision of
over the global $\pr(\kappax)$. For any given line of sight the
convergence estimate can be made before the investment of expensive follow-up
observations. We find that the lowest $\kappax$ lines of sight have the best constrained 
$\pr(\kappax)$ after applying our reconstruction. We find that selecting systems by the precision with which
$\kappax$ can be estimated leads to a sample of unbiased
time delay distance estimates with 1\% uncertainties due to external convergence effects.
Photometric data alone is sufficient to pre-select the best lenses and can be done
before investment in lightcurve monitoring. Preferentially selecting lines of
sight with high external shear (as might happen in an assembled sample of
quadruply-imaged time delay lenses) produces time-delay distance estimates with
typical uncertainties of 1.8\% and biases of 0.8\%. We suggest areas for the improvement of this general
analysis framework that should allow yet more accurate cosmological inferences
to be made.

\end{abstract}

@@

The \MS results have been used to analyse real observations as well. 
\citet{SuyuEtal2010} selected \MS lines of sight by their apparent
galaxy over-density, in a 45 arcsec radius aperture down to $i < 24.5$,
to match the observed over-density towards the time delay lens
B1608$+$656 \citep{FassnachtEtal2011}.  The resulting distribution of
$\kappax$ values from the ray-tracing was taken to be an estimate of 
$\Pr(\kappax)$, which was then marginalised over when inferring the time
delay distance in this system. A similar procedure was used in \citet{SuyuetAl2012b}.
In a companion paper to this one, Greene et al. (submitted) investigate
improvements to this method by weighting the galaxy counts by galaxy
mass, and perpendicular distance to the line of sight, again using the
\MS mock catalogues and their associated $\kappax$ values to construct
$\Pr(\kappax|\data)$. For the most over-dense lines of sight the
Green et al. results  show a $\sim$20\% improvement over
number counts alone, but little improvement for less dense lines.

@@

\new{
In strong lens systems where the source is time variable, the images do not 
vary simultaneously; the optical path length for each image is different due
to relativistic and geometric effects. The difference in optical path length
causes a time-delay between the light curves of each image. \citeauthor{FalcoEtal1985}
showed that the presence of additional matter (in the form of a mass-sheet)
along the line of sight has no observable effect except to rescale the time-delays.
As such, it is necessary to include $\kappax$ in the lens modelling, if
cosmological parameters are to be estimated accurately and precisely from
observed time delays \citep{SuyuEtal2010}. 
}

@@

Mass distributions that are
physically associated with the lens galaxy will affect the stellar dynamics of
the lens galaxy, so that dynamical observations can break the internal
mass-sheet degeneracy by providing an additional estimate of the lens galaxy's
mass \citep[e.g.,][]{Koopmans2004;SuyuEtal2010}. In this work our aim is to quantify the
lensing effect of massive objects, regardless of their position along the line
of sight.

@@

Transforming an observed photometric galaxy catalogue into a 
catalogue of halo masses, positions, and redshifts, will enable us to
attempt a reconstruction of the convergence induced by every halo near a
line of sight. \new{We also need to account for the convergence due to dark structures and
the divergence due to voids; we do not model voids or dark structures, but their
effects are included statistically by calibration the the convergence in halos,
$\kappah$, against the ray-traced convergence, $\kappax$, for \MS lines of sight.}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{figs/cornerplot.eps}
\caption[Biased?]{The joint distribution of 
$\kappah$ and $\kappax$, from 100,000 reconstructed lines of
sight. $\kappah$ traces $\kappax$, but with a positive offset which arises from
$\kappah$ not accounting for any voids, dark structures or smooth mass
component. At fixed $\kappah$ the scatter in $\kappax$ grows with $\kappah$.}
\label{fig:jointkh-k}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{figs/globaldist.eps}
\caption[magcut]{Recovering the global $\pr(\kappax)$, shown in red, 
from $10^{5}$ ray-traced lines of
sight. The black curve shows a histogram of 
sample reconstructed 
$\kappax$ values, each one drawn from an inferred PDF 
$P(\kappax|\{\Mhalo,z\}$)
in one of $10^{5}$ reconstructed lines of sight. 
The reconstructions were performed given
perfect knowledge of halo mass and redshift -- in this case
the reconstruction recovers the correct convergence.}
\label{fig:globaldist}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{figs/Width.eps}
\caption{The widths of the \infered PDFs $\pr(\kappah|\data)$ for
$10^5$ lines of sight, given different quality of data. Blue:
spectroscopic redshift for every halo with $i<26$; Purple: 
spectroscopic redshift for every halo with $i<23$, and every halo with $i<24$
within 1 arcminute; Cyan: spectroscopic redshift for every halo with $i<22$; 
Green: all the objects in the field have only photometric redshifts.}
\label{fig:reconwidths}
\end{figure}

\begin{figure}
\includegraphics[width=\columnwidth]{figs/Widthvshilberrt.eps}
\caption{Width of the \infered PDF $\pr(\kappax|\data)$ versus the true
$\kappax$ along a line of sight. Blue assumes a spectroscopic redshift for
every halo, whilst green assumes only photometric redshifts. The width of the
\infered $\pr(\kappah|\data)$ increases with $\kappax$.}

@@

\begin{figure*}
\includegraphics[width=\textwidth]{figs/biasplots.eps}
\caption{Convergence accuracy from combining subsets of
lenses. By subtracting the true (ray-traced) convergence from the
\infered $\pr(\kappax)$ for each line of sight we are left with a PDF that
should be estimating zero; these can be multiplied together for multiple
lines of sight to emulate a joint likelihood analysis, and 
test for possible bias in it. These plots show the
expectation value of $\prod_{i=1}^N \pr_i(\kappax-\kappax^{\rm true})$ --
deviations from zero constitute biases. The solid, dashed and dotted
lines correspond to combinations of 
20, 5 and 2 sightlines respectively. Green lines show the results
from multiplying $\pr(\kappax|\data)$ \infered from photometry alone, whilst red
lines are from multiplying $\pr(\kappax)$ \infered from photometry and a
spectroscopic redshift for all halos with $i<23$ and all halos within 1
arcminute and $i<24$. The leftmost plot shows randomly selected lines
of sight. The central plot shows sight lines randomly selected from the
33\% of lines whose P($\kappax$) is most tightly constrained. The right
hand plot shows only lines of sight with external shear of 0.03 or
greater.}
\label{fig:biasplots}
\end{figure*}

>>>>

High precision cosmological distance measurements towards individual objects
such as time delay gravitational lenses or type Ia supernovae are affected by
weak lensing perturbations by galaxies and groups along the line of sight.
In time delay gravitational lenses,
``external convergence,'' $\kappax$, can dominate the uncertainty in the inferred distances.
In this paper we attempt to reconstruct $\kappax$,
due to line of sight structure using a simple halo model.
 We use mock catalogues from the Millennium Simulation, and compare
our reconstructed $\pr(\kappax|\data)$ to ray-traced $\kappax$ truth values: 
taking into account realistic uncertainties on redshift and stellar masses, we
typically find that the reconstruction yields a factor of 2 improvement in precision of
over the global $\pr(\kappax)$. We find that the lowest $\kappax$ lines of sight have the best constrained 
$\pr(\kappax)$ after applying our reconstruction. We find that selecting systems by the precision with which
$\kappax$ can be estimated leads to a sample of unbiased
time delay distance estimates each with 1\% uncertainties due to external convergence effects.
Photometric data alone is sufficient to pre-select the best lenses and can be done
before investment in lightcurve monitoring. Preferentially selecting lines of
sight with high external shear (as might happen in an assembled sample of
quadruply-imaged time delay lenses) produces time-delay distance estimates with
typical external convergence uncertainties of 1.8\% but with systematic errors of 0.8\%.
We suggest areas for the improvement of this general analysis framework that
should allow yet more accurate cosmological inferences to be made.
\end{abstract}

@@

The \MS results have been used to analyse real observations as well. 
\citet{SuyuEtal2010} selected \MS lines of sight by their apparent
galaxy over-density, in a 45 arcsec radius aperture down to $i < 24.5$,
to match the observed over-density towards the time delay lens
B1608$+$656 \citep{FassnachtEtal2011}.  The resulting distribution of
$\kappax$ values from the ray-tracing was taken to be an estimate of 
$\Pr(\kappax)$, which was then marginalised over when inferring the time
delay distance in this system. A similar procedure was used in \citet{SuyuEtal2012}.
In a companion paper to this one, Greene et al. (submitted) investigate
improvements to this method by weighting the galaxy counts by galaxy
mass, and perpendicular distance to the line of sight, again using the
\MS mock catalogues and their associated $\kappax$ values to construct
$\Pr(\kappax|\data)$. For the most over-dense lines of sight the
Green et al. results  show a $\sim$20\% improvement over
number counts alone, but little improvement for less dense lines.

@@

In strong lens systems where the source is time variable, the images do not 
vary simultaneously; the optical path length for each image is different due
to relativistic and geometric effects. The difference in optical path length
causes a time-delay between the light curves of each image. \citeauthor{FalcoEtal1985}
showed that the presence of additional matter (in the form of a mass-sheet)
along the line of sight has no observable effect except to rescale the time-delays.
As such, it is necessary to include $\kappax$ in the lens modelling, if
cosmological parameters are to be estimated accurately and precisely from
observed time delays \citep{SuyuEtal2010}.

@@

Mass distributions that are
physically associated with the lens galaxy will affect the stellar dynamics of
the lens galaxy, so that dynamical observations can break the internal
mass-sheet degeneracy by providing an additional estimate of the lens galaxy's
mass \citep[e.g.,][]{Koopmans2004,SuyuEtal2010}. In this work our aim is to quantify the
lensing effect of massive objects, regardless of their position along the line
of sight.

@@

Transforming an observed photometric galaxy catalogue into a 
catalogue of halo masses, positions, and redshifts, will enable us to
attempt a reconstruction of the convergence induced by every halo near a
line of sight. {We also need to account for the convergence due to dark structures and
the divergence due to voids; we do not model voids or dark structures, but their
effects are included statistically by calibration the the convergence in halos,
$\kappah$, against the ray-traced convergence, $\kappax$, for \MS lines of sight.}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{figs/cornerplot.eps}
\caption[Biased?]{The conditional distribution of 
$\kappax$ given $\kappah$, from $10^5$ reconstructed lines of
sight. $\kappah$ was generated from perfect knowledge of halo mass and redshift. Solid (dashed) lines show the 68\% (95\%) confidence interval $\kappah$ traces $\kappax$, but with a positive offset which arises from
$\kappah$ not accounting for any voids or dark structures.
 At fixed $\kappah$ the scatter in $\kappax$ grows rapidly with $\kappah$.}
\label{fig:jointkh-k}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{figs/globaldist.eps}
\caption[magcut]{Recovering the global $\pr(\kappax)$, shown in red, 
from $10^{5}$ ray-traced lines of
sight. The black curve shows a histogram of 
sample reconstructed 
$\kappax$ values, each one drawn from an inferred PDF 
$P(\kappax|\kappah)$ in one of $10^{5}$ reconstructed lines of sight. 
The reconstructions were performed given
perfect knowledge of halo mass and redshift -- in this case
the reconstruction recovers the correct convergence. The reconstruction assumes
perfect knowledge of the halo mass and redshift.}
\label{fig:globaldist}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{figs/Width.eps}
\caption{The widths of the \infered PDFs $\pr(\kappah|\data)$ for
$10^5$ lines of sight, given different quality of data. Blue:
spectroscopic redshift for every halo with $i<26$; Red: 
spectroscopic redshift for every halo with $i<23$, and every halo with $i<24$
within 1 arcminute;
Green: all the objects in the field have only photometric redshifts.}
\label{fig:reconwidths}
\end{figure}

\begin{figure}
\includegraphics[width=\columnwidth]{figs/Widthvshilberrt.eps}
\caption{Width of the \infered PDF $\pr(\kappax|\data)$ versus the true
$\kappax$ along a line of sight. Green assumes only photometric redshifts
whilst red assumes a campaign of targetted spectroscopy. The width of the
\infered $\pr(\kappah|\data)$ increases with $\kappax$.}

@@

\begin{figure*}
\includegraphics[width=\textwidth]{figs/biasplots.eps}
\caption{Convergence accuracy from combining subsets of
lenses. By subtracting the true (ray-traced) convergence from the
\infered $\pr(\kappax)$ for each line of sight we are left with a PDF that
should be estimating zero; these can be multiplied together for multiple
lines of sight to emulate a joint likelihood analysis, and 
test for possible bias in it. These plots show the
expectation value of $\prod_{i=1}^N \pr_i(\kappax-\kappax^{\rm true})$ --
deviations from zero constitute biases. The solid, dashed and dotted
lines correspond to combinations of 
20, 5 and 2 sightlines respectively. Green lines show the results
from multiplying $\pr(\kappax|\data)$ \infered from photometry alone, whilst red
lines are from multiplying $\pr(\kappax)$ \infered from photometry and a
spectroscopic redshift for all halos with $i<23$ and all halos within 1
arcminute and $i<24$. The leftmost plot shows randomly selected lines
of sight. The central plot shows sight lines randomly selected from the
33\% of lines whose P($\kappax$) is most tightly constrained by our model. The right
hand plot shows only lines of sight with external shear of 0.05 or
greater.}
\label{fig:biasplots}
\end{figure*}
